

```{r}
setwd("C:/Users/Eric/Desktop/Reproducible research")
storm <- read.csv("C:/Users/Eric/Desktop/Reproducible research/repdata%2Fdata%2FStormData.csv.bz2", sep = ",", header= TRUE)
str(storm)
```

```{r}

length(unique(storm$EVTYPE))
```

## Changing Event type to lower cas
```{r}
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))

storm$EVTYPE <- event_types

library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
```
## Events that causes death and injury
```{r}
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)

##Top 10 events that caused largest number of deaths are:
  
  fatal_events[, c("EVTYPE", "fatalities")]
```

## Top 10 events that caused largest number of deaths are:


```{r}  
  injury_events[, c("EVTYPE", "injuries")]
```


Calculating cost associated with property and crop damage:

```{r}
exp_transform <- function(e) {
  # h -> hundred, k -> thousand, m -> million, b -> billion
  if (e %in% c('h', 'H'))
    return(2)
  else if (e %in% c('k', 'K'))
    return(3)
  else if (e %in% c('m', 'M'))
    return(6)
  else if (e %in% c('b', 'B'))
    return(9)
  else if (!is.na(as.numeric(e))) # if a digit
    return(as.numeric(e))
  else if (e %in% c('', '-', '?', '+'))
    return(0)
  else {
    stop("Invalid exponent value.")
  }
}

prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```

## Compute the economic loss by event type

```{r}
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

## filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```
 
 ## Top 10 events that caused most property damage (in dollars) are as follows
```{r}  
  prop_dmg_events[, c("EVTYPE", "prop_dmg")]
  
  ## Similarly, the events that caused biggest crop damage are
  
  crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```  
  
  Health Impact -The following plot shows top dangerous weather event types.
```{r}  
library(ggplot2)

p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")
p1
p2
``` 
  Conclusions
Tornadoes causes the higest number of deaths and injuries compared to all the other disastrous events followed by wind floods and excessive heat.
  
  
##Economic impact of weather events and associated cost since 1950
  
``` {r} 
 ## Set the levels in order
  
p1 <- ggplot(data=prop_dmg_events,
               aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")
  
  p2 <- ggplot(data=crop_dmg_events,
               aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")
p1
p2
  
```  
##Conclusion
The results shows that flash floods and thunderstorm causes the largest property damage among the different natural disaster.

  